Development of a educational and sustainable robotic platform for children and adults

### 2014-05-03 doc/paper.tex
stuff

\documentclass[11pt,a4paper]{article}
%\documentclass[conference]{IEEEtran}
\usepackage[english]{babel} 
\usepackage[T1]{fontenc}
\usepackage{pdfpages}
\usepackage{lmodern}
\usepackage[utf8]{inputenc}
\usepackage{eurosym}
\DeclareUnicodeCharacter{20AC}{\euro}
\usepackage{csquotes}
\usepackage{graphicx}
\usepackage{rotating} 
\usepackage{array}
\usepackage{booktabs} 
\usepackage{longtable}
\usepackage{ifthen}
\usepackage{xcolor}
\usepackage{tabu}
\usepackage{colortbl}
\usepackage{calc}
\usepackage{pifont}
\usepackage{forloop}
\usepackage[nomessages]{fp}
\usepackage{caption}
\usepackage{subcaption}

@@

@@

\subsection{Arduino Robot}
\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{images/arduinorobot.jpg}
  \caption{Picture of the Arduino Robot}
\end{figure}

@@

\subsection{LEGO Mindstorms NXT 2.0}
\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{images/mindstorms.jpg}
  \caption{Picture of the LEGO Mindstorms NXT (from http://www.devoxx.com/)}
\end{figure}

@@

\subsection{Thymio II}
\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{images/thymioii.jpg}
  \caption{Picture of the Thymio II (from https://aseba.wikidot.com/en:thymio)}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_ballcaster.jpg}
  \caption{Metal ball caster (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_dcmotor.jpg}
  \caption{Standard DC motor}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_gearboxmotor.jpg}
  \caption{Gearbox motor with wheel}
  \end{subfigure}
  \caption{Different DC motor types (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_speedsensor.jpg}
  \caption{Speedsensor (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{table}[h!]
\centering
\begin{tabular}{p{0.15\textwidth}p{0.15\textwidth}p{0.5\textwidth}}
\toprule
Pin count & Pin type & Description \\
\midrule
2 & PWM  & Motor speed control using pulse-width-modulation.\\
4 & GPIO & Direction control and steering control using a 4-bit (16 combinations) pattern.\\
\bottomrule
\end{tabular}
\caption{DC motor microcontroller pin usage}
\label{tbl:dc_pin}
\end{table}

@@

\begin{table}[h!]
\centering
\begin{tabular}{p{0.15\textwidth}p{0.15\textwidth}p{0.5\textwidth}}
\toprule
Pin count & Pin type & Description \\
\midrule
1 & GPIO & Clock which clocks the serial data. Not time critical due to the design of a shift register.\\
1 & GPIO & Serial data in which sets the bit pattern.\\
1 & GPIO & Chip enable which enables the shift register output.\\
\bottomrule
\end{tabular}
\caption{Shift register microcontroller pin usage}
\label{tbl:74hc595_pin}
\end{table}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{images/mecanum.png}
  \caption{Screenshot of a mecanum wheel 3D model (from http://www.thingiverse.com/thing:2318)}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{images/omniwheel.jpg}
  \caption{Control scheme of a omni wheel (from http://commons.wikimedia.org/wiki/File:Omni\_wheel\_control.jpg)}
  \label{fig:omnicontrol}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_ir.jpg}
  \caption{IR Reciever with integrated Schmitt trigger (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_proximity_front.jpg}
  \caption{front-view}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_proximity_back.jpg}
  \caption{side-view}
  \end{subfigure}
  \caption{IR proximity sensor (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_ultrasonic.jpg}
  \caption{HC-SR04 ultrasonic distance sensor (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_compass.jpg}
  \caption{Triple axis compass on breakout-board (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_irdiodes.jpg}
  \caption{Infrared diode (sender)}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_ir.jpg}
  \caption{Infrared receiver (reciever)}
  \end{subfigure}
  \caption{Components for infrared communication (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_nrf24l01.jpg}
  \caption{nRF24L01 breakout board (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_lithium-polymer.jpg}
  \caption{Lithium-polymer battery (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\textwidth]{images/schematic/stepup.png}
  \caption{Stepup converter schematic using the MCP1650}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=\textwidth]{images/schematic/charge.png}
  \caption{Lithium-polymer battery charge circuit using the MCP73831}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_tftfront.jpg}
  \caption{front-view}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_tftback.jpg}
  \caption{back-view}
  \end{subfigure}
  \caption{1.8 Inch TFT display (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.7\textwidth]{images/part_types_thtsmt.jpg}
  \caption{Illustration of THT (top) and SMT (bottom) parts on a PCB (green) with Solder (grey) (own illustration)}
\end{figure}

@@

Based on these guidelines the PCB is created. At first the components of the Arduino Leonardo compatible unit (first CPU/user CPU) are layouted. Special care is taken with the spacing between the Arduino shield compatible pin headers.\footnote{See the simplified board layout in the Appendix for illustration.}

>>>>

\documentclass[11pt,a4paper]{article}
%\documentclass[conference]{IEEEtran}
\usepackage[english]{babel} 
\usepackage[T1]{fontenc}
\usepackage{pdfpages}
\usepackage{lmodern}
\usepackage[utf8]{inputenc}
\usepackage{eurosym}
\DeclareUnicodeCharacter{20AC}{\euro}
\usepackage{csquotes}
\usepackage{graphicx}
\usepackage{rotating} 
\usepackage{array}
\usepackage{booktabs} 
\usepackage{longtable}
\usepackage{ifthen}
\usepackage{xcolor}
\usepackage{tabu}
\usepackage{colortbl}
\usepackage{calc}
\usepackage{pifont}
\usepackage{forloop}
\usepackage[nomessages]{fp}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}

@@

\begin{abstract}
Kinder, Jugendliche und Erwachsene an Technik heranzuführen ist eine wichtige Aufgabe in einer zunehmend technisierten Gesellschaft. Ein verbreiteter Ansatz ist die Heranführung mit Hilfe von Robotern da diese ihre Umgebung wahrnehmen und mit dieser interagieren können.

Eine Evaluation zeigt, dass viele Robotikplattformen für den Bildungssektor Kriterien an eine nachhaltige und pädagogisch wertvolle Lösung zu Teilen bereits erfüllen. Kriterien wie günstige Anschaffungskosten, Erweiterbarkeit und geringe Komplexität sind nur zu Teilen erfüllt und zeigen die Notwendigkeit einer neuen Plattform.

Nach einer detalierten Analyse zu Bewegungskonzepten, Sensoren, Kommunikation, Mikrocontroller-Plattformen und Batterie-Systemen wird anhand einer Auswahl von Komponenten eine kostengünstige, pädagogisch wertvolle, quelloffene und erweiterbare Robotikplattform konstruiert. Eine detalierte Darstellung der Entwicklungsschritte zeigt auf, welche Designentscheidung die Bewertungskriterien positiv beeinflussen.

Die Abschließende Evaluation von Chancen und Risiken zeigt, dass eine pädagogisch sinnvolle und wirtschaftliche Verwendung der neu entwickelten Robotikplattform möglich ist. Die zuvor auf andere Plattformen angewendeten Evaluationskriterien zeigen eine deutliche Verbesserung in Punkten Erweiterbarkeit, Anschaffungskosten und Nachhaltigkeit.
\end{abstract}

\begin{abstract}
Getting children, young adults and adults in touch with technology is important with today’s technologically advancing society. A common approach is to use Robots since they can sense their environment and modify it.

The evaluation of different robotic platforms designed for educational use shows that criteria such as sustainability and educational value are met to a certain extend. Other criteria such as low cost of purchase, extendibility and low complexity are partially met and therefore show the need for a new robotic platform.

After a detailed analysis of movement concepts, sensors, communication techniques, microcontroller platforms and power supply systems the optimal components for the new  platforms were chosen. The choice is based on cost, educational, extension and open source criteria. A detailed description of the construction steps allows to comprehend design decisions which positively influence the evaluation criteria.

A evaluation of chances and challenges on the constructed robot shows that a educational use of the platform is possible whilst operating economically. The previously used evaluation criteria for evaluating other platforms applied to the new platform shows enhancements in extendibility, purchase costs and sustainability.
\end{abstract}

@@

\subsection{Arduino Robot}
\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{images/arduinorobot.jpg}
  \caption{Picture of the Arduino Robot}
\end{figure}

@@

\subsection{LEGO Mindstorms NXT 2.0}
\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{images/mindstorms.jpg}
  \caption{Picture of the LEGO Mindstorms NXT (from http://www.devoxx.com/)}
\end{figure}

@@

\subsection{Thymio II}
\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{images/thymioii.jpg}
  \caption{Picture of the Thymio II (from https://aseba.wikidot.com/en:thymio)}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_ballcaster.jpg}
  \caption{Metal ball caster (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_dcmotor.jpg}
  \caption{Standard DC motor}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_gearboxmotor.jpg}
  \caption{Gearbox motor with wheel}
  \end{subfigure}
  \caption{Different DC motor types (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_speedsensor.jpg}
  \caption{Speedsensor (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{table}[H]
\centering
\begin{tabular}{p{0.15\textwidth}p{0.15\textwidth}p{0.5\textwidth}}
\toprule
Pin count & Pin type & Description \\
\midrule
2 & PWM  & Motor speed control using pulse-width-modulation.\\
4 & GPIO & Direction control and steering control using a 4-bit (16 combinations) pattern.\\
\bottomrule
\end{tabular}
\caption{DC motor microcontroller pin usage}
\label{tbl:dc_pin}
\end{table}

@@

\begin{table}[H]
\centering
\begin{tabular}{p{0.15\textwidth}p{0.15\textwidth}p{0.5\textwidth}}
\toprule
Pin count & Pin type & Description \\
\midrule
1 & GPIO & Clock which clocks the serial data. Not time critical due to the design of a shift register.\\
1 & GPIO & Serial data in which sets the bit pattern.\\
1 & GPIO & Chip enable which enables the shift register output.\\
\bottomrule
\end{tabular}
\caption{Shift register microcontroller pin usage}
\label{tbl:74hc595_pin}
\end{table}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{images/mecanum.png}
  \caption{Screenshot of a mecanum wheel 3D model (from http://www.thingiverse.com/thing:2318)}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{images/omniwheel.jpg}
  \caption{Control scheme of a omni wheel (from http://commons.wikimedia.org/wiki/File:Omni\_wheel\_control.jpg)}
  \label{fig:omnicontrol}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_ir.jpg}
  \caption{IR Reciever with integrated Schmitt trigger (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_proximity_front.jpg}
  \caption{front-view}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_proximity_back.jpg}
  \caption{side-view}
  \end{subfigure}
  \caption{IR proximity sensor (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_ultrasonic.jpg}
  \caption{HC-SR04 ultrasonic distance sensor (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_compass.jpg}
  \caption{Triple axis compass on breakout-board (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_irdiodes.jpg}
  \caption{Infrared diode (sender)}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_ir.jpg}
  \caption{Infrared receiver (reciever)}
  \end{subfigure}
  \caption{Components for infrared communication (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_nrf24l01.jpg}
  \caption{nRF24L01 breakout board (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\textwidth]{images/30_lithium-polymer.jpg}
  \caption{Lithium-polymer battery (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{images/schematic/stepup.png}
  \caption{Stepup converter schematic using the MCP1650}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{images/schematic/charge.png}
  \caption{Lithium-polymer battery charge circuit using the MCP73831}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_tftfront.jpg}
  \caption{front-view}
  \end{subfigure}
  \begin{subfigure}{0.48\textwidth}
  \centering
  \includegraphics[width=0.9\linewidth]{images/30_tftback.jpg}
  \caption{back-view}
  \end{subfigure}
  \caption{1.8 Inch TFT display (CC-BY-SA Ben Oswald)}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.7\textwidth]{images/part_types_thtsmt.jpg}
  \caption{Illustration of THT (top) and SMT (bottom) parts on a PCB (green) with Solder (grey) (own illustration)}
\end{figure}

@@

Based on these guidelines the PCB is created. At first the components of the Arduino Leonardo compatible unit (first CPU/user CPU) are layouted. Special care is taken with the spacing between the Arduino shield compatible pin headers. The USB connector for the IC is placed near the edge of the board to allow a good fit for the cable.

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{robot_schematic_user.pdf}
  \caption{Robot schematic creation. Arduino Leonardo compatible circuit part is higlighted in dark green.}
\end{figure}

The next step is to place logically connected components together. The battery charging circuit components are placed near the edge for a good placement of the batteries JST connector. Placing the connector in the middle of the board makes it difficult to carry the battery.

When routing power circuits it is important to set the right trace width. A trace with 10 mil width on 35 micrometer copper can carry up to 1A but heats up 30 degree kelvin. A 20 mil trace only heats up 10 degree kelvin when carrying the same load. \cite[p. 841]{horowitz1989art}

In the robotic platform developed the trace width for power lines is 16 mil. The expected heat up is tolerable since in normal operation the average current should be much lower.

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{robot_schematic_battery.pdf}
  \caption{Robot schematic creation. Battery charging circuit components are higlighted in dark red.}
\end{figure}

Followed by the step-up circuit parts most of power related components are placed on the board. The components are placed near the battery charging circuit to reduce the length of the circuits. All components are placed with enough spacing to make soldering easier. This specially holds for the MCP1650 since the leads have a spacing of 0.65mm only. Making this chip easy accessible for desoldering wick helps to re-position it in case of misplacement.

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{robot_schematic_stepup.pdf}
  \caption{Robot schematic creation. Step-Up circuit parts are higlighted in dark blue.}
\end{figure}

The 3.3V LDO is mainly for providing power to the optional TFT display, the NRF24L01 module and shields which harvest energy from the 3.3V pins at the shield connector. To reduce the loss of energy due to trace-length the components are placed near the mentioned parts. To stabilize the current flow additional capacitors are placed on the output pins.

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{robot_schematic_3_3v.pdf}
  \caption{Robot schematic creation. 3.3V LDO components are higlighted in orange.}
\end{figure}

Having all power related components placed on the board, the second Arduino Leonardo compatible circuit for controlling sensors, actuators and optional modules is placed on the board.

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{robot_schematic_brain.pdf}
  \caption{Robot schematic creation. Second Arduino Leonardo compatible circuit components are higlighted in purple.}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{robot_schematic_motor.pdf}
  \caption{Robot schematic creation. Motor control circuit components are higlighted in lime.}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{robot_schematic_sensor.pdf}
  \caption{Robot schematic creation. Line sensing components are higlighted in brown.}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{robot_schematic_addtional.pdf}
  \caption{Robot schematic creation. Additional components are higlighted in aqua.}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{robot_schematic_optional.pdf}
  \caption{Robot schematic creation. Placeholder for optional components on breakout boards are higlighted in olive.}
\end{figure}
